Strict Pure Functional Programming and Automatic Parallelization\\ (Project-in-progess Paper)

### 2016-02-14 pubs/tfp16/paper.tex
more words

% This is LLNCS.DEM the demonstration file of
% the LaTeX macro package from Springer-Verlag
% for Lecture Notes in Computer Science,
% version 2.4 for LaTeX2e as of 16. April 2010
%
\documentclass{llncs}
%
\usepackage{makeidx}  % allows for indexgeneration
%
\begin{document}
%
%mkd \frontmatter          % for the preliminaries
%
\pagestyle{headings}  % switches on printing of running heads
%
%mkd \mainmatter              % start of the contributions
%
\title{Strict Pure Functional Programming and Automatic Parallelization}
%
\titlerunning{Automatic Parallelization}  % abbreviated title (for running head)
%                                     also used for the TOC unless
%                                     \toctitle is used
%
\author{Kei Davis\inst{1} \and Dean Prichard\inst{1} \and David Ringo\inst{1,2}}
%
\authorrunning{Kei Davis et al.} % abbreviated author list (for running head)
%
%%%% list of authors for the TOC (use if author list has to be modified)
%\tocauthor{Ivar Ekeland, Roger Temam, Jeffrey Dean, David Grove,
%Craig Chambers, Kim B. Bruce, and Elisa Bertino}
%
\institute{Los Alamos National Laboratory, Los Alamos, NM, USA\\
\email{kei.davis@lanl.gov},\\ WWW home page:
\texttt{http://ccsweb.lanl.gov/\homedir kei/}
\and
University of New Mexico,
Albuquerque, NM, USA}

@@

\begin{abstract}
The trend is functional programming in scientific computing, and in particular
strict (by default) functional programming, in various guises.  The project is
the demonstration of a light-weight, higher-order, polymorphic,
strict-by-default purely functional language implementation in which we can
experiment with automatic parallelization strategies and varying degrees of
default strictess.

@@

\section{Background and Motivation}

\section{Goals}

@@

\section{Current Status and Expected Results (by the time of the symposium)}

@@

>>>>

% This is LLNCS.DEM the demonstration file of
% the LaTeX macro package from Springer-Verlag
% for Lecture Notes in Computer Science,
% version 2.4 for LaTeX2e as of 16. April 2010
%
\documentclass{llncs}
%
\usepackage{makeidx}  % allows for indexgeneration
%
\begin{document}
%
%mkd \frontmatter          % for the preliminaries
%
\pagestyle{headings}  % switches on printing of running heads
%
%mkd \mainmatter              % start of the contributions
%
\title{Strict Pure Functional Programming and Automatic Parallelization\\
       (Project-in-progess Paper)}
%
\titlerunning{Automatic Parallelization}  % abbreviated title (for running head)
%                                     also used for the TOC unless
%                                     \toctitle is used
%
\author{Kei Davis\inst{1} \and 
        Dean Prichard\inst{1} \and 
        David Ringo\inst{1,2} \and
        Loren Anderson\inst{1,3} \and
        Jacob Marks\inst{1,4}\thanks{All contributors, past and present, are listed.}
}
%
\authorrunning{Kei Davis et al.} % abbreviated author list (for running head)
%
%%%% list of authors for the TOC (use if author list has to be modified)
%\tocauthor{Ivar Ekeland, Roger Temam, Jeffrey Dean, David Grove,
%Craig Chambers, Kim B. Bruce, and Elisa Bertino}
%
\institute{Los Alamos National Laboratory, Los Alamos, NM, USA\\
\email{kei.davis@lanl.gov},\\ WWW home page:
\texttt{http://ccsweb.lanl.gov/\homedir kei/}
\and
University of New Mexico,
Albuquerque, NM, USA
\and
North Dakota State University,
Fargo, North Dakota, USA
\and 
New Mexico Institute of Mining and Technology,
Socorro, New Mexico, USA
}

@@

% moronic author listing code increments footnote counter
\setcounter{footnote}{0}

\begin{abstract}
The trend is functional programming in scientific computing, and in particular
strict (by default) functional programming, in various guises.  Our project is
the demonstration of a light-weight, higher-order, polymorphic pure functional
language implementation in which we can experiment with automatic
parallelization strategies and varying degrees of default strictness.
Strictness prescribes the degree of potential parallelism, that is, we are not
exploring speculative evaluation.

\emph{Preliminary paper note: We are reporting work in progress.  At the time
  of writing we have a complete serial implementation and are currently
  developing the parallel runtime.  We expect to have preliminary results from
  our parallel implementation by the time of the symposium.}

@@

\section{Background and Motivation}
The practice of highest-performance scientific computing (HPC) is, overall,
highly conservative with respect to change.  Even as the US government commits
to a push for exascale computing ($10^{18}$ FLOPS) within the next decade, a
highly respected member of the HPC community recently stated ``Fortran is
essential for exascale programming.''~\cite{Heroux16}.  Existing HPC codes run
to hundreds of thousands of lines and complete rewrites are infeasible.  C++
has become more common for new code starts, but their nature remains the same:
huge codes with poor (or non-existent) encapsulation of side effects, and
multiple levels of parallelism also poorly abstracted, especially at the
thread (shared-memory) level.  Thread-level parallel efficiency can be poor
and is explained by Amdahl's law because the parallel model is bulk synchronous.

\subsection{The trend: strict but still pure functional programming}
There is a small but growing understanding that \emph{pure functional
  semantics}, in some form, is essential to reining in the complexity of these
ever-evolving scientific codes.  Such semantics are expressed in a number
of ways as highlighted following.

\subsubsection{Rely on programmer discipline.}
Legacy Fortran scientific codes likely represent worst practices in state
encapsulation, with most data global and accessible to all.  It is almost
ironic, then, that of the mainstream high-performance scientific programming
languages (Fortran, C, and C++), Fortran 95 was the first to introduce the
\texttt{pure} function qualifier, and with enforcement by the compiler.
Otherwise it is entirely up to the programmer to make functions composable
and thread safe.  Here at LANL this approach is being used to good effect
to clean up some legacy code bases.

\subsubsection{Obey constraints prescribed by a parallel runtime system.}
There are a number of parallel runtime systems that encourage (but cannot
enforce) a pure functional style.  We will highlight Stanford's Legion
runtime as an example~\cite{Bauer12}.

In Legion, non-function-local data, or \emph{regions}, are requested from the
runtime system.  Regions may be shared among \emph{tasks} (essentially C++
functions), and each task registers its access (read, write, read/write, etc.)
and coherency requirements for each region to which it will have access.
Other than registered access to regions tasks must not access non-constant
global data.  In serial execution these access requirements are notionally
superfluous, but in parallel execution the runtime can dynamically calculate
the dependency graph and relax the programmatically defined strict ordering of
task execution.  Very high parallel efficiency has been demonstrated on extremely
large computing systems~\cite{Bauer14}.  Here at LANL the Legion system is
being strongly considered for a new large-scale C++ code effort.\footnote{
It is interesting to note that the PI of the Legion project was previously a
co-developer of a strict, arguably pure functional language
implementation~\cite{AikenFL}.}

\subsubsection{Use a pure functional language.}
Some twenty years after the US Department of Energy laboratory complex was
presented with the most performant and highly parallel implementation of a
pure functional language to date~\cite{Davis96}, and collectively had no
conceptual understanding of what it was, much less why it might be useful or
important, there is a small number of laboratory HPC practioners who are
developing an appreciation for pure functional programming in Haskell.  Still,
they see their needs as different than other Haskell users: performance is
paramount, unboxed arrays are a (perhaps \emph{the}) primary data structure,
and laziness is more of a nuisance--even a serious hindrance--than a
convenience, much less algorithmically essential in the sense of Bird's
\emph{repmin}~\cite{Bird84} or Johnsson's general attribute
grammars~\cite{Johnsson87}.

A common complaint among these budding high-performance functional programmers
is the need to get strictness just right: to remove space leaks,
improve performance, get communication (e.g.\ MPI) and IO to work as they
expect, and so on.  More specifically, they expect to be able to reason about
space and time usage using their existing mental models. While they are
happy to write code that is strongly reminiscent of the numerical examples in
Hughes' classic piece (whether they've read it or not)~\cite{Hughes89},
they would also be happy to write in a slightly modified style that does
not require non-strictness---after all, that's how they've been thinking their
entire careers.

Much anecdotal evidence suggests that many others have similar difficulties
with lazy (or non-strict) by default semantics, with the most telling recent
concrete evidence being the push for, and realization of, the \texttt{Strict}
and \texttt{StrictData} language extensions newly available in GHC 8.0.
Quoting the documentation:
\begin{quote}
High-performance Haskell code (e.g. numeric code) can sometimes be littered
with bang patterns, making it harder to read. The reason is that laziness
isn't the right default in this particular code, but the programmer has no way
to say that except by repeatedly adding bang patterns~\cite{strict-strictdata}.
\end{quote}

We claim, then, that there is a trend towards \emph{strict-by-default pure
  functional programming}, whether as imposed by programmer discipline, a parallel
runtime system (such as the aforementioned Legion, wherein the pure semantics
at function granularity allows the runtime system to relax the nominal serial
ordering), or the language itself (e.g.\ strict Haskell).

Intel's TBB?  Cilk++?

\subsection{Strictness and Parallelism}

It is a trivial observation that given strict functional semantics, wherein
function arguments may be safely evaluated before function evaluation, it is
safe to evaluate the arguments, and the function, in parallel, and that this
principle applies recursively in the (dynamic) expression tree (or graph).

\section{Project and Goals}

Our project is the light-weight implementation of a pure, higher-order,
polymorphic, functional language and runtime system with which we can
experiment with automatic parallelization strategies with degrees of language
strictness ranging from lazy to fully strict.  Light-weight has several
meanings or implications.  First and foremost it means that \emph{we are not
  attempting to compete with ghc in any respect}.\footnote{We estimate that we
  will have, to within an order of magnitude, one person-hour of available
  development time for each person-year that ghc currently represents, or put
  another way, one second for each hour, respectively.}

\begin{itemize}
\item The implementation must be feasible in terms of effort;
\item The language must be strongly typed, implicitly or explicitly.  More
  specifically, the language should explicitly support both Hindley-Milner
  polymorphism and unboxed types~\cite{firstclass};
\item Straighforward direct linkage with C, preferably without a sophisticated
  foreign function interface, is essential;
\item The language should be ``currying-friendly'' as described in the ``fast
  curry'' paper~\cite{fastcurry};
\item Choice of degree of strictness is easily selected;
\item Sharing, in the sense of what ``lazy'' means vs.\ merely non-strict,
  should be preserved for all degrees of strictness;
\item Tail calling should be properly implemented, i.e., not as a trampoline;
\item etc.
\end{itemize}

For feasibility, we restrict ourselves to shared-memory implicit parallelism,
but distributed-memory parallelism should be straightforwardly realizable
using e.g.\ MPI or POSIX sockets.

\subsection{Varying degrees of strictness}

First some words about terminology.  For all degrees of strictness sharing (as
implied by lazy but not by non-strict) of heap-allocated objects is preserved
we will not use the term \emph{lazy} and simply refer to strictness
properties.  As is common we will use strict in an operational sense, meaning
that argument(s) are evaluated before being passed in a function call.
Similarly, when referring to language semantics we refer to the operational
rather than denotational semantics.

Greater strictness implies more opportunities for parallelism, possible losses
or gains in space efficiency, and less expressiveness.  We seek to explore
the interplay of all of these.

ghc 8.0 gives three possibilities: conventional non-strict, constructor
strict, and function- and constructor strict, but there are possible
variations.  The obvious variation is function, but not constructor, strict.
In Haskell the distinction between constructors and functions is somewhat
blurred because constructors behave like functions (outside of pattern
matching), but this need not be the case.  In STG---ghc's intermediate
language---the distinction is clear: heap objects are explicitly constructed
by \texttt{let} constructs in which constructor names are explicitly mentioned
(not aliased); constructor names are not functions.

Other possible variations concern the notional arity of the underlying
function being applied.  This is readily explained in terms of the operation of the
STG machine described next.

In STG the arity of a function is syntactically defined as the number of
manifest formal parameters, thus \texttt{id x = x} has by definition arity 1
regardless of what type it might be dynamically ``instantiated'' at.  We distinguish
\emph{application}, e.g.\ \texttt{f x y}, which is a syntactic notion, from
actually calling a function. Thus \texttt{id id x} is an application of \texttt{id}
to two arguments, but the underlying \emph{FUN} (user- or system- defined function)
when actually called is given exactly one argument.

In the eval-apply model, given application \texttt{f x y}, in the standard
(non-strict) semantics \texttt{f} is first evaluated and the arity of the
underlying \emph{FUN} (user- or system- defined function) or \emph{PAP}
(partial application of a \emph{FUN}) is determined---in the case of
\emph{PAP} the arity is that of the underlying \emph{FUN} less the number of
arguments already provided.  If the arity is greater than two then a new
\emph{PAP} is created with arguments \texttt{x} and \texttt{y} inserted and
the new \emph{PAP} returned.  If the arity is equal to two then the underlying
\emph{FUN} is tail-called with any previous argument (in case of \emph{PAP})
and new arguments \texttt{x} and \texttt{y}.  If the arity is less than two
(it must be at least one, so one in this case) the underlying \emph{FUN} is
called-with-return with any previous argument (in case of \emph{PAP}) and new
argument \texttt{x}, then the object returned applied to remaining argument
\texttt{y}.

Given application \texttt{f x y} when might \texttt{x} and \texttt{y} be
evaluated?  In the non-strict semantics evaluation is on demand, after
\texttt{f} has been evaluated, the underlying \emph{FUN} actually called
(i.e., not just partially applied and wrapped up in a \emph{PAP}), and one of
those arguments forced in subsequent evaluation (assuming that they weren't
forced in the evaluation of \texttt{f} itself).

In a strict semantics one can come up with various evaluation orders, but
excluding the absurd, as far as distinguishing termination properties goes
these can be distilled down to two possibilities, namely whether on
construction of a \emph{PAP} the (new) arguments are forced or not.  (An
absurd strategy might force every second argument.)  These are in fact
distinguishable because (in Haskell) the first argument to \texttt{seq} might
evaluate to a \emph{PAP}.

This then gives three reasonable operational modes that are strictly ordered
in terms of termination properties: non-strict, fully strict---all arguments
in an application are evaluated---and the intermediate in which only those
arguments needed by a function call are evaluated.

-- spell out STG syntax?

@@

Our wish list sounds much like the ghc Haskell implementation, and indeed ghc
is part of our master plan.  In short, our plan is to use ghc as our front-end
to generate its intermediate form, STG, then escape to our system from there.
There are various possible ways to accomplish the former: using ghc's option
to dump STG as text (see Stuart's report~\cite{Stuart}), hacking ghc itself
and, optionally, using ghc as a library.  We note that this approach has 
a precedent in the aforementioned Nvidia research compiler.

\subsection{STG to C}

The back-end is written in Haskell and generates proper tail-calling C code;
the runtime system is written in C/C++.  The system fairly faithfully
implements the STG machine as described in Peyton-Jones and Marlow's ``fast
curry'' paper~\cite{fastcurry} and dynamically illustrated by Pope's
miniSTG~\cite{ministg} with the exception of the state transition rules which,
of course, are different than those for lazy semantics.  Sharing is preserved.

\subsection{Strictness}

The STG machinery makes the choice of evaluation strategy easy to change, both
in terms of function and constructor application.

\section{Current Status, and Expected Results by the time of the symposium}

No bridge from ghc STG to our STG yet.

At the time of writing we have a complete serial implementation:  STG parser, type inferencer, code
generator, and runtime system including a garbage collector.

C stack, our stack.

Mutual exclusion for garbage collector--heap check with every heap allocation.  In fact garbage collector
can be invoked at any point in time except when populating a heap object immediately after its allocation.

\subsection{Mini-Haskell}

\subsection{Type system}

We perform standard Hindley-Milner type inference and enforcement, augmented
with built-in and user-defined unboxed types as described by Peyton-Jones and
Launchbury's seminal work~\cite{unboxed}, but for unboxed types code
generation is currently limited to built-in single-word (64-bit) or smaller
types (e.g., \texttt{Int\#}, \texttt{UInt\#}, \texttt{Float\#}, \texttt{Double\#})
and user-defined enumerations, e.g., \texttt{data unboxed Bool\# = False\# |
  True\#}.

\subsection{Test suite and testing}

Extensive prelude.

Test suite of several hundred small programs.

Students learning Haskell and STG, writing Mini-Haskell and STG programs that
serendipitously go into our automated testing suite which is partitioned
according to our lattice of evaluation strategies (3 degrees of function
strictness by two degrees of constructor strictness), which tells us how much
non-strictness each program needs and something about the style in which
programs are written.

Continuous build system (Jenkins)

\subsection{Parallel evaluation}

Parallel runtime is being implemented following the ideas of~\cite{SPJs-many-papers}, thread pool,
stack and heap for each thread, blah, blah.

\section{Other Related Work}
Nvidia research compiler.

\section{Summary, Future Work, Conclusions?}

@@

\bibitem{Heroux}
January 14, 2016
Exascale Programming: Adapting What We Have Can (and Must) Work
Michael A. Heroux, Sandia National Laboratories
http://www.hpcwire.com/2016/01/14/24151/

\bibitem{Davis96}
MPP Parallel Haskell. 
K. Davis. 
Draft Proceedings Implementation of Functional Languages 1996 (IFL '96) LNCS
1268, pp49-54, Bonn/Bad-Godesberg, Germany, LNCS 1268, Springer-Verlag.

\bibitem{Bauer12} 
Legion: Expressing Locality and Independence with Logical
  Regions. M. Bauer, S. Treichler, E. Slaughter and A. Aiken. In Proceedings
  of the Conference on Supercomputing, pages 1-11, November 2012.

\bibitem{Bauer14} 
Structure Slicing: Extending Logical Regions with Fields. M. Bauer, S. Treichler, E. Slaughter and A. Aiken. In Proceedings of the Conference on Supercomputing, pages 845-856, November 2014.

\bibitem{AikenFL}
A. Aiken, J.H. Williams, and E.L. Wimmers
The FL Project:  The Design of a Functional Language
(technical report not published elsewhere), September 1993.
http://theory.stanford.edu/~aiken/publications/trs/FLProject.pdf

\bibitem{Bird84}
R. S. Bird, 
Using circular programs to eliminate multiple traversals of data, 
Acta Informatica, 21(3), 239-250 (1984).

\bibitem{Johnsson87}
Thomas Johnsson
Attribute Grammars as a Functional Programming Paradigm
Proc. of a conference on Functional programming languages and computer architecture
Pages 154-173 
Springer-Verlag London, UK ©1987 
table of contents ISBN:0-387-18317-5

\bibitem{Hughes89}
J. Hughes,
Why Functional Programming Matters,
Computer Journal,
32,
2,
98--107,
1989.

\bibitem{strict-strictdata}
http://ghc.haskell.org/trac/ghc/wiki/StrictPragma

[ghclib] GHC/As a library http://wiki.haskell.org/GHC/As\_a\_library
